\DocumentMetadata{
  pdfversion=2.0,
  pdfstandard=a-4,
  lang=en-GB,
}

\documentclass[letter, nobiblatex]{modern-doc}

\setkomavar{fromname}{Oscar Wilde}
\setkomavar{fromaddress}{
  H.M. Prison Reading\\
  Reading, Berkshire\\
  England%
}
\setkomavar{signature}{Oscar Wilde}
\setkomavar{place}{Reading Gaol}
\setkomavar{date}{April 1st, 1897}

\begin{document}

\begin{letter}{
  Robert Ross, Esq.\\
  London\\
  England%
}

\setkomavar{subject}{Correspondence regarding Manuscript and Literary Matters}

\opening{My dear Robbie,}

I send you a MS. separate from this, which I hope will arrive safely. As soon as
you have read it, I want you to have it carefully copied for me.

Luckily I suffer from headaches when I read my Greek and Roman poets---so they
have not been of much use---but his kindness was great in getting the set. Ask
him to express my gratitude to the lady who lives at Wimbledon. Write to me,
please, in answer to this and tell me about literature, what new books, etc.,
also Jones' play and Forbes-Robertson's management---about any new tendency in
the stage of Paris or London.

Also try and see what Lemaitre, Bauer, and Sarcey said of \q{Salome,} and give me
a little resume: please write to Henri and say I am touched at his writing
nicely. Robert Sherard knows him. It was sweet of you to come and see me. You
must come again next time. Here I have the horror of death with the still
greater horror of living, and in silence and misery.

I always remember you with deep affection. I wish Ernest would get from Oakley
Street my portmanteau, fur-coat, clothes, and the books of my own writing which I
gave my dear mother---ask in whose name the burial ground of my mother was
taken.

\closing{Always your friend,}

\encl{Manuscript (separate)}

\end{letter}

\end{document}